---
title: "NYC flights "
format: html
editor: visual
---

## Libraries

```{r}
library(DBI)
library(RSQLite)
library(nycflights13)
library(magrittr)
library(ggplot2)

```

## Create Connection

```{r}
# create connection to the local database 
con <- dbConnect(SQLite(), "nycflights.sqlite")

# load the tables into the database
dbWriteTable(con, "flights",  nycflights13::flights,  overwrite = TRUE)
dbWriteTable(con, "airlines", nycflights13::airlines, overwrite = TRUE)
dbWriteTable(con, "airports", nycflights13::airports, overwrite = TRUE)
dbWriteTable(con, "planes",   nycflights13::planes,   overwrite = TRUE)
dbWriteTable(con, "weather",  nycflights13::weather,  overwrite = TRUE)

```

## Research Question

What factors are associated with departure delays for flights departing from New York City airports in 2013?

## Exploratory Data Analysis

This exploratory data analysis investigates the associations between departure delays and different airports, flight times, airlines, destinations and weather conditions. These descriptive results do not establish causal effects or test statistical significance; they identify patterns for follow up analysis.

**Airports:** Did flights from EWR, JFK, or LGA have different average departure delays?

```{r}
 origin_plot =  dbGetQuery(con, "
             SELECT 
             origin AS Origin,
             AVG(dep_delay) AS avg_dp_delay
             FROM flights 
             GROUP BY origin 
             ORDER BY AVG(dep_delay) DESC
             ") %>%
   ggplot(mapping = aes(x = reorder(Origin, avg_dp_delay), y = avg_dp_delay)) +
   geom_col() +
   labs(title = "Average Departure Delay by Origin",
        x = "Origin Airport",
        y = "Average Departure Delay (minutes)") + 
   theme_bw() +
   geom_text(
     aes(label = round(avg_dp_delay, 1)),
     vjust = -0.4,
     size = 4) +
   expand_limits(y = 16)
 
 origin_plot
```

The three New York City airports did have differing average departure delays with EWR having the highest at 15 minutes, followed by JFK's 12 minutes and finally LGA's 10 minutes.

**Timing:** How did delays vary by month and scheduled departure hour?

```{r}
dbGetQuery(con, "
           SELECT 
           month, 
           AVG(dep_delay) AS avg_dp_delay
           FROM flights 
           GROUP BY month 
           ORDER BY AVG(dep_delay) DESC
           ")

```

Departure delay does appear to vary by month. June and July have the highest average departure delays of around 21 and 22 minutes respectively and December has the third highest with 17 minutes.

```{r}
hour_plot = dbGetQuery(con, "
           SELECT 
           hour AS departure_hour, 
           AVG(dep_delay) AS avg_dep_delay, 
           COUNT(*) AS number_of_flights
           FROM flights 
           WHERE hour IS NOT NULL AND dep_delay IS NOT NULL
           GROUP BY hour 
           ORDER BY AVG(dep_delay) DESC
           ") %>% 
  ggplot(mapping = aes(x = departure_hour, y = avg_dep_delay)) +
  geom_line() +
  theme_bw() +
  labs(title = "Average Departure Delay by Scheduled Hour", 
       y = "Average Departure Delay (minutes)",
       x = "Scheduled Departure Hour") +
  theme_bw()

hour_plot
```

It does look like average departure delay is highest around 8:00 pm and then tapers off towards both midnight and 5am. Additionally, most flights appear to be in between the hours of 3 pm to 8 pm with another spike from 6 am to around 8 am.

**Airline:** Which airlines had the highest delays? To make this a fair comparison, we will only look at airlines with more than 10,000 flights to reduce smaller airlines with less flights affecting our analysis.

```{r}
dbGetQuery(con, "
           SELECT 
           a.name AS airline,
           AVG(f.dep_delay) AS avg_dp_delay 
           FROM flights f 
           JOIN airlines a 
           ON f.carrier = a.carrier 
           GROUP BY a.name 
           HAVING COUNT(*) >= 10000 
           ORDER BY AVG(f.dep_delay) DESC
           ")
```

Different airlines do appear to have different average delays. The three airlines with the highest average departure delays are ExpressJet with an average of 20 minutes, Southwest with an average of 18 minutes and Endeavor Air with an average of 17 minutes.

**Route:** Which destinations had the highest observed average departure delay?

```{r}
dbGetQuery(con, "
           SELECT 
           dest AS destination, 
           AVG(dep_delay) AS avg_dep_delay,
           COUNT(dep_delay) AS flights
           FROM flights 
           GROUP BY dest 
           HAVING COUNT(dep_delay) > 100
           ORDER BY AVG(dep_delay) DESC
           LIMIT 5
           ")
```

It appears that CAE, TUL and OKC had the highest average departure delays with 36, 35 and 31 minutes respectively.

**Weather:** Were visibility, precipitation, wind, or temperature associated with departure delays?

```{r}
weather_delays = dbGetQuery(con, "
                            SELECT 
                            f.flight,
                            f.dep_delay, 
                            w.wind_speed, 
                            w.precip, 
                            w.visib, 
                            w.temp
                            FROM flights f
                            JOIN weather w 
                            ON f.origin = w.origin 
                            AND f.time_hour = w.time_hour
                            WHERE f.dep_delay IS NOT NULL
                            AND w.wind_speed IS NOT NULL
                            AND w.precip IS NOT NULL
                            AND w.visib IS NOT NULL
                            AND w.temp IS NOT NULL
                            ")
```

```{r}
weather_delays %>%
  ggplot(mapping = aes(x = visib, y = dep_delay)) +
  geom_bin2d() +
  labs(title = "Visibility and Departure Delay", 
       x = "Visibility",
       y = "Departure Delay (minutes)") +
  theme_bw()
```

It appears that there is not any strong association between visibility and departure delay.

```{r}
weather_delays %>%
  ggplot(mapping = aes(x = precip, y = dep_delay)) +
  geom_bin2d() +
  labs(title = "Precipitation and Departure Delay", 
       x = "Precipitation (inches)",
       y = "Departure Delay (minutes)") +
  theme_bw()
  
```

The scatter plot is dominated by flights where there was no precipitation making it irresponsible to draw conclusions from this plot.

```{r}
weather_delays %>%
  ggplot(mapping = aes(x = wind_speed, y = dep_delay)) +
  geom_bin2d() +
  labs(title = "Wind Speed and Departure Delay", 
       x = "Wind Speed (mph)",
       y = "Departure Delay (minutes)") +
  theme_bw()
```

There does not appear to be any strong association between wind speed and departure delay.

```{r}
weather_delays %>%
  ggplot(mapping = aes(x = temp, y = dep_delay)) +
  geom_bin2d() +
  labs(title = "Temperature and Departure Delay", 
       x = "Temperature (F)",
       y = "Departure Delay (minutes)") +
  theme_bw()
```

There does not appear to be any strong association between temperature and Departure Delay.

Because departure delays vary between EWR, LGA and JFK, later analysis could determine whether the above factors differ within the three airports.

## Disconnect

```{r}
dbDisconnect(con) 
```

## Save Plots

```{r}
ggsave(
  "output/figures/delay_by_origin.png",
  origin_plot,
  width = 7,
  height = 5,
  dpi = 300
)

ggsave(
  "output/figures/delay_by_hour.png",
  hour_plot,
  width = 7,
  height = 5,
  dpi = 300
)


```
